---
title: "Peck - Final Capstone Project"
author: "Jennifer Peck"
date: "2023-01-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

For my final project in the Data Science certification course, I decided to use
a data set that was pertaining to student loans. Not only is this pertinent to 
me as a student myself, but it is also the industry I have worked in for the 
past 20 years. On Kaggle, I was able to find a data set that is a scrape of 
COD (Common Origination and Disbursement) for the 2020-2021 academic school year.
COD is a website used by the Department of Education, student loan servicers, 
schools, and students to track loan information. 

This data set is focused to school level items. There is information about what 
type of school each is, where it is located, the number of recipients of loans, 
as well as the number of loans and dollar amounts for both originations and 
disbursements. For this project, I am going to focus on originations as that is 
the picture of the total loan debt taken out. I am going to view and present the
data by modeling different methods of query and graphing, and determine if I 
locate any trends. This trending data may be on type of school, state location, 
or loan type. I will present the results in multiple formats, utilizing both 
tables and graphs. Let's take a look!

## Methods and Analysis

### Set-Up

To get started, I need to download the data set into R, and get it into a format
that will make it easier to use going forward. I also want to make sure I have 
all the various library packages installed and loaded. The file on Kaggle is 
called "The schools that create the most student debt", which I am renaming to
simply "debt". 

```{r, echo=FALSE}
#Loading needed packages

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(readr)) install.packages("readr", repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(ggrepel)) install.packages("ggrepel", repos = "http://cran.us.r-project.org")
library(tidyverse)
library(caret)
library(readr)
library(dplyr)
library(ggplot2)
library(ggrepel)

#Kaggle data set "The schools that create the most student debt"
#https://www.kaggle.com/datasets/thedevastator/the-schools-that-create-the-most-student-debt
#https://github.com/jenpeck/Capstone_Final_Project.git

options(timeout = 120)
dl <- "Student Loan Debt by School 2020-2021.csv"
if(!file.exists(dl))
  download.file("https://github.com/jenpeck/Capstone_Final_Project/blob/main/
                Student%20Loan%20Debt%20by%20School%202020-2021.csv", dl)


#this takes the csv and gives it an easier alias
debt <- read.csv("Student Loan Debt by School 2020-2021.csv")
```

I want to take a quick look at what we are starting with. First, I determine 
the class type of my file is a data frame, and I look at an example of the 
structure. I see that there are 42765 objects and 13 variables.

```{r, echo=FALSE}
#This looks at the class type and what the data frame looks like
class(debt)
str(debt)
```


I decided to rename all of the columns as some of them have special characters
that I don't want to deal with.

```{r}
#The original csv has special characters in some of the column names so I want
#to rename the columns for ease going forward
names(debt)[1] <- 'index'
names(debt)[2] <- 'opeid'
names(debt)[3] <- 'school'
names(debt)[4] <- 'city'
names(debt)[5] <- 'state'
names(debt)[6] <- 'zip'
names(debt)[7] <- 'school_type'
names(debt)[8] <- 'loan_type'
names(debt)[9] <- 'num_recipients'
names(debt)[10] <- 'num_loans_originated'
names(debt)[11] <- 'dollars_originated'
names(debt)[12] <- 'num_disbursements'
names(debt)[13] <- 'dollars_disbursed'

#check to make sure the columns are renamed
colnames(debt)
```



I want to view a tibble of this data frame to see if it is getting into shape. 
It's starting to look pretty good!

```{r, echo=FALSE}
#get a look at the table
debt %>% as_tibble()
```
\newpage
Now I want to start getting an idea of some of my populations. Based on this 
summary, I have 4078 schools, 7 loan types, and 8 school types.

```{r, echo=FALSE}
#look at a summary of some of the table items. 
table1 <- debt %>% summarize(distinct_schools = n_distinct(school), 
                  distinct_loantype = n_distinct(loan_type),
                  distinct_schooltype = n_distinct(school_type))
knitr::kable(table1 ,caption = "Summary")
```

Let's check out the loan types: 

```{r, echo=FALSE}
#what are the 7 loan types and counts of how many rows each 
table2 <- debt %>% group_by(loan_type) %>% summarize(count = n()) %>% 
  arrange(desc(count))
knitr::kable(table2 ,caption = "Loan Types and #Rows")
```

This brings up a few interesting points. I see there are 25 rows that have no 
loan type, the count of the labeled loan types are all equal, and there is a 
huge number under 'total'. I will deal with the blanks in a bit and I assume the 
number of each loan type is even because there is a line item for each type, 
whether the school reported any info there or not. First though, I am curious
about what the 'totals' are. I am going to narrow down to looking at only the 
info for one school to try and figure it out. 

```{r}
debt %>% filter(school == 'UNIVERSITY OF ALABAMA')
```

It appears that one of the total lines is the sum of the values 
of the other loan types added together. I cannot determine what the other 4 
total lines are showing. To have it not skew my results going forward, I am 
going to throw out all of the rows in the set with a loan type of 'total'. I am 
also going to throw out the 25 loans with no loan type. I will call this new 
result 'clean_debt'. 

```{r}
clean_debt <- debt %>% filter(loan_type != 'Total')
clean_debt <- clean_debt %>% filter(loan_type != "")

#make sure that I no longer show either totals or blanks
clean_debt %>% group_by(loan_type) %>% summarize(count = n())

#check to see how far row count went down
nrow(clean_debt)
```

When I ran the first tibble of this data, it jumped out at me that one of the 
lines showing had 0's reporting for a row - indicating that they reported their
school's information to COD, but didn't happen to have any loans of a particular 
type. Even though 0's won't add to my results, I am going to toss those lines as 
well, just to make our data set as clean as possible. 

```{r, echo=FALSE}
#throw out rows with 0 recipients
clean_debt <- clean_debt %>% filter(num_recipients != '0')
```

Finally, in terms of set up, I discovered later in my research that there are 
schools listed in the data that are not located in America. Because I intend to
look at trending by state, I want to remove these lines as well. An example of 
this is shown with this school in Ireland. 

```{r, echo=FALSE}
#found this later and then decided to bring in at the beginning. Found there
#were schools outside of usa. I want to remove those since I am going to be 
#looking at some of the info by state. By running this example below I found 
#that the international schools have a blank state so that is how I will filter 
#them out
clean_debt %>% filter(school == 'ROYAL COLLEGE OF SURGEONS IN IRELAND')
clean_debt <- clean_debt %>% filter(state != '')
```

```{r}
#how many rows left
nrow(clean_debt)
```

This gets the cleaned up data down to 13283 lines. The data is now in good 
shape and cleaned up of items that will give skewed results. 


### Data Modeling

Ok, I am ready to start looking for trends! I will split this section by looking
for trends from four angles - school type trends, loan type trends, state trends, 
and school trends.

#### School Type  

The first thing I want to check out is how many schools there are for each 
school type. I will view this via bar graph. 

```{r, echo=FALSE, fig.align='center'}
#how many schools of each school type - graphed. Need to first make a temp table
#so that I can narrow down to distinct schools rather than rows

distinctschools <- clean_debt %>% select(school, school_type) %>%  
  distinct(school, school_type)

distinctschools %>% 
  ggplot(aes(school_type, fill = school_type)) + 
  geom_bar(color = "black") + 
  theme_gray() +
  ggtitle("Distibution of School Types") +
  xlab("School Types") +
  ylab("Count of Schools")
```

The graph above shows 3 school types: Private-Nonprofit, Proprietary, and Public.
Private-Nonprofit has the most schools reporting. If you recall, before we 
cleaned up the data, we originally had 8 school types showing. The other 5 
that are no longer here were indicating either foreign schools, blanks, or 
invalid information (info in wrong column). 

Next, I am going to see how many loan recipients there are at each school type. 
The results show that there are close to double the amount of recipients at 
public schools vs private schools, even though there are more private schools on
the list, as we saw above. 

```{r, echo = FALSE, fig.align='center'}
#how many recipients in each school type
table3 <- clean_debt %>% group_by(school_type) %>% 
  summarize(number_recipients = sum(num_recipients)) %>% 
  arrange(desc(number_recipients))
knitr::kable(table3 ,caption = "Recipients at School Types")
```

For the last portion of my review by loan type, I want to look at the amount of 
dollars originated, by each. Here we see that the most money is in the realm of 
public schools, but it's a lot closer to the private school amount than the
number of recipients would lead us to believe. 

```{r, echo = FALSE, fig.align='center'}
#how much money originated by school type
table4 <- clean_debt %>% group_by(school_type) %>% 
  summarize(dollars_orig = sum(dollars_originated)) %>% 
  arrange(desc(dollars_orig))
knitr::kable(table4 ,caption = "Total Orig by School Type")
```


#### Loan Type  

The second angle I want to look into is trends by loan type. 

In the graph below, we see that the amounts of Subsidized and Unsubsidized - 
Undergraduate loans are fairly close. This is what I would expect to see as it 
is common for students to get both a subsidized and unsubsidized loan at the 
same time. 

```{r, echo=FALSE, fig.align='center'}
#graph loans by loan type. first have to make a temp table to sum the number of
#loans
sumlnorig <- clean_debt %>% select(loan_type, num_loans_originated) %>%  
  group_by(loan_type) %>%
  summarize(sum_ln_orig=sum(num_loans_originated))

sumlnorig %>% 
  ggplot(aes(x=loan_type, y=sum_ln_orig)) + 
  geom_bar(stat="identity", fill = "skyblue", color = "black", ) + 
  geom_text(aes(label = signif(sum_ln_orig)), nudge_y = 100000) +
  theme_gray() +
  ggtitle("Distibution of Loan Types") +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Types of Loans") +
  ylab("Count of Loans")
```

I want to see how many loan recipients there are by each loan type. If we 
compare this to the graph above, the counts are very close but there are more
loans by type than there are recipients. This makes sense as some students get
more than one loan. What we can't tell here is distinct recipient. If one 
student got both a subsidized and unsubsidized loan, they will show up as a 
recipient for each column. 

```{r, echo = FALSE, fig.align='center'}
#how many recipients have each loan type on the clean debt data set
table5 <- clean_debt %>% group_by(loan_type) %>% 
  summarize(number_recipients = sum(num_recipients)) %>%
  arrange(desc(number_recipients))
knitr::kable(table5 ,caption = "Recipients and Loan Types")
```

Next, let's check out the amount of money originated by each loan type. 
Unsubsidized - Graduate is the highest dollar amount. This is likely due to the
cost of Grad school being higher than Undergrad.

```{r, echo = FALSE, fig.align='center'}
#how much money originated by loan type
table6 <- clean_debt %>% group_by(loan_type) %>% 
  summarize(dollars_orig = sum(dollars_originated)) %>% 
  arrange(desc(dollars_orig))
knitr::kable(table6 ,caption = "Orig by Loan Type")
```

For the final look at loan types, I want to determine what the average dollar 
amount originated by recipient is. The results show that on average, recipients 
that take out Grad PLUS loans have the highest average loan amount. Subsidized 
is the lowest. 

```{r, echo = FALSE, fig.align='center'}
#figure out average $ originated by recipient by loan type
table7 <- clean_debt %>% group_by(loan_type) %>% 
  summarize(avgorig = sum(dollars_originated)/sum(num_recipients)) %>%
  arrange(desc(avgorig))
knitr::kable(table7 ,caption = "Average Orig by Person and Loan Type")
```


#### State Information  

The third angle that I would like to investigate is at the state level. I am 
curious if there are trends of how many loans or amount of money that is 
originated by state. 

First up, I'd like to find out which 10 states have the most schools reporting 
in this data set. California has over 1300 schools, while the next closest state
is New York with 859.

```{r, echo=FALSE, fig.align='center'}
#how many total schools per state - top 10 states
table8 <- clean_debt %>% count(state) %>% top_n(10, n) %>% arrange(desc(n))
knitr::kable(table8 ,caption = "Top 10 Schools per State")
```


I queried how many schools, per school type, are being reported in each state. 
This is a full look, so the table is long. 

```{r, echo = FALSE, fig.align='center'}
#how many schools by type per state
table9 <- clean_debt %>% count(state, school_type)
knitr::kable(table9 ,caption = "Schools, by Type, per State")
```

Now I want to look at the 10 states that have the highest number of loan 
recipients Texas is the highest with almost 532 thousand. 

```{r, echo = FALSE, fig.align='center'}
#top 10 most recipients by state, descending
table10 <- clean_debt %>% group_by(state) %>% 
  summarize(number_recipients = sum(num_recipients)) %>%
  top_n(10, number_recipients) %>% arrange(desc(number_recipients))
knitr::kable(table10 ,caption = "Top 10 Recipients by State")
```

Here is a larger overview of recipients by state in a descending graph view.

```{r, echo=FALSE, fig.align='center'}
#graph recipients by state desc
clean_debt %>% group_by(state) %>% 
  summarize(number_recipients = sum(num_recipients)) %>% 
  mutate(state = reorder(state, number_recipients)) %>%
  ggplot(aes(state, number_recipients)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  ggtitle("Recipients by State") +
  xlab("")
```

Next, I want to view the dollars originated by state. Here are the top 10 states
with the highest originated amount. California lands in the top spot. 

```{r, echo = FALSE, fig.align='center'}
#figure out sum of $ originated by state - show top 10
table11 <- clean_debt %>% group_by(state) %>% 
  summarize(dollars_orig = sum(dollars_originated)) %>%
  top_n(10, dollars_orig) %>% arrange(desc(dollars_orig))
knitr::kable(table11 ,caption = "Dollars Orig per State")
```
 
Similar to before, I want to look at a descending graph by state, this time with
a full view of total originated amounts for all states.

```{r, echo=FALSE, fig.align='center'}
#graph originations by state desc
clean_debt %>% group_by(state) %>% 
  summarize(dollars_orig = sum(dollars_originated)) %>% 
  mutate(state = reorder(state, dollars_orig)) %>%
  ggplot(aes(state, dollars_orig)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  ggtitle("Originations by State") +
  xlab("")
```

I want step away from totals for a bit and look at the average amount originated 
by recipient, by state. Interestingly, Washington DC and Puerto Rico borrow the 
highest amount per person. 

```{r, echo = FALSE, fig.align='center'}
#figure out average $ originated by recipient by state.
table12 <- clean_debt %>% group_by(state) %>% 
  summarize(avgorig = sum(dollars_originated)/sum(num_recipients))  %>%
  top_n(10, avgorig) %>% arrange(desc(avgorig))
knitr::kable(table12 ,caption = "Average Orig, by Recipient, by State")
```

I decided to graph out this data in a different way. This makes it a little 
easier to see that PR and DC are outliers at the high end of dollars per person,
Guam is an outlier on the low end of the same thing. TX, CA, PA, and NY are 
outliers in terms of the amount of people who got loans, but their average dollar
amount falls in line with most other states. 

```{r, echo=FALSE, fig.align='center'}
#graph to show where the states land when you look at number of recipients by 
#average dollars originated. Need to make a temp table first for the calculations
thing <- clean_debt %>% 
  group_by(state) %>% 
  summarize(sumrec = sum(num_recipients), 
            sumorig = sum(dollars_originated),
            avgorig = sum(dollars_originated)/sum(num_recipients))              

thing %>% 
  ggplot(aes(sumrec, avgorig, label = state)) +   
  geom_point() +
  geom_text_repel() + 
  xlab("Number of Recipients") + 
  ylab("Avg Dollars Originated") +
  ggtitle("Avg Dollars Orig by Recipient, by State") 
```


#### School Information  

In the last section of data analysis, I am going to look at some trends by 
individual schools. If we look at the schools with the most money originated, 
Arizona State University landed on top. This is interesting because Arizona 
hasn't popped up as being at the top of any of our other queries. 

```{r, echo = FALSE, fig.align='center'}
#figure out most $ originated by school - show top 10
table13 <- clean_debt %>% group_by(school) %>% 
  summarize(dollars_orig = sum(dollars_originated))  %>%
  top_n(10, dollars_orig) %>% arrange(desc(dollars_orig))
knitr::kable(table13 ,caption = "Top 10 Most Orig by School")
```

I also want to see which schools have the highest average amount originated by
recipient. An interesting discovery here is that the top 10 are all either 
medical schools or schools best known for their medical programs. 

```{r, echo = FALSE, fig.align='center'}
#figure out school with highest avg $ by recipient. 
table14 <- clean_debt %>% group_by(school) %>% 
  summarize(avgorig = sum(dollars_originated)/sum(num_recipients))  %>%
  top_n(10, avgorig) %>% arrange(desc(avgorig))
knitr::kable(table14 ,caption = "Schools with Highest Avg Orig by Recipient")
```


## Results

Looking at this student loan debt by school information in multiple ways has 
shown a number of interesting results. 

 - Far more people obtain the assistance of student loans if they attend public 
schools. This may indicate a couple of things - lower income people are more 
likely to attend public school, or more people in general attend public school. 
We can't determine this without data regarding total number of attendees. 

 - Even though there was close to double the amount of recipients at public vs 
private schools, the dollar amount originated was fairly close. This tells us
that the private school students need to borrow more money, which makes sense 
as private schools tend to have far more expensive tuition. 

 - Looking at the data by loan type doesn't help to draw very many conclusions, 
but there was one interesting result - graduate students need to borrow far
more per person than anybody else. Unsubsidized-Graduate and Grad PLUS loans
average out to $20-25,000 per recipient, while Unsubsidized-Undergrad and 
Subsidized loans come out to roughly $4100 per recipient. 

 - Trending by state showed that the most schools are in CA, followed by NY and 
TX, however the state with the most students getting a loan is TX, followed by 
CA. NY was 4th on that list. 7 of the top 10 states by number of recipients are 
also on the list of top 10 states with the most amount of schools. The top 10
states by recipients match the top 10 states by dollars originated. This likely
tells us that these states aren't outliers, but simply at the top due to volume.

 - When the state data was graphed by average originated dollars per recipient, 
it was easier to find some outliers. Puerto Rico and Washington DC have a higher
proportion of money originated per person than the rest of the group. Guam was 
at the other end of the spectrum, with very few recipients and originated loans.

 - Arizona State University was the school that originated the most amount of 
debt. This is interesting as Arizona as a state didn't come up anywhere else 
while we were looking at trends. It isn't in the top 10 for overall originations, 
number of recipients or anything else, yet the school on its own originated over
a half billion dollars in one academic year. My assumption is if we had numbers 
of total school attendees, we would find that a higher proportion of students 
who attend ASU borrow money than other colleges. 

 - The final query regarding which schools have the highest average of dollars 
originated by recipient was interesting, though not entirely surprising. All of 
the top 10 schools were either medical schools, or colleges that were best known
for their medical programs. 

## Conclusion

Based on this data set, if you are a student obtaining a graduate degree or a 
medical degree, and need to borrow money in order to do it, you are going to be
taking on a large debt. Undergraduate degrees and public schools, are likely to 
cause less of a financial strain, though they may add up if you obtain multiple
loans. Since this data was a snapshot of one academic year, we are unable to see
the average amount borrowed by recipient over the length of their college degree. 
The academic year for this data was also 2020-2021 so COVID would play a part in
decreasing how many people went to school and needed to obtain loans. I would 
expect that if we looked at this same data from 2018-2019, the numbers would be
even larger. 

## References

The schools that create the most student debt
Data set - Kaggle
https://www.kaggle.com/datasets/thedevastator/the-schools-that-create-the-most-
student-debt?resource=download

Introduction to Data Science: Data Analysis and Prediciton Algorithms with R
Rafael A. Irizarry
http://rafalab.dfci.harvard.edu/dsbook/

R Markdown Cookbook
Yihui Xie, Christophe Dervieux, Emily Riederer
https://bookdown.org/yihui/rmarkdown-cookbook/

Github data visualization cheat sheet
https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf

Github R markdown cheat sheet
https://github.com/rstudio/cheatsheets/blob/main/rmarkdown.pdf

My github for this project
https://github.com/jenpeck/Capstone_Final_Project



